Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-02-09 Paper/Draft_harmonization_paper.tex
09.02.: Brooke edits Pt.5

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Overview_sim_results.png}
  \caption{Ranges of aggregated trait affinities for the three examples of taxonomic hierarchies and simulated levels of trait variability. Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $*$ of the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots.}
  \label{fig:overview_sim_results}
\end{figure}

@@

\subsection*{Trait definition discrepancies and taxonomic resolution}

Synthesizing trait information from multiple trait databases is crucial for developing the full potential of trait-based approaches, e.g. for studying community trait responses to environmental gradients worldwide. Our attempt at harmonising invertebrate trait information from trait databases of different regions showed that harmonisation was labour-intensive because grouping features are resolved differently into traits and various codings are used to describe traits in the various databases. In addition, the same traits are sometimes defined differently, requiring expert knowledge to reclassify these traits (e.g. the trait piercer in the Tachet database). Although, others have noted the lack of standardised trait terminology in freshwater ecology (\cite{baird_toward_2011, brink_traits-based_2011}), we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features. To resolve definition discrepancies and facilitate data synthesis in the future, terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organism groups, e.g. for plants with the \textit{Thesaurus Of Plant characteristics} (TOP) initiative (\cite{garnier_towards_2017}). The core of this harmonisation initiative is to provide standardised trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by \citet{baird_toward_2011} a decade ago. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS), providers of invertebrate trait data could connect their traits to such ontologies (\cite{schneider_towards_2019}). Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and ultimately interpretation of the derived results.

Our analysis aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets. Although many taxa are covered in the trait databases from which we extracted our datasets, the availability of trait information varies strongly across grouping features and taxonomic resolution differs between databases. Much information is available for grouping features that are often used in trait-environment studies, such as feeding mode and respiration. Surprisingly, there is little information on body form, a trait that is relatively easy to determine.

While trait information from Europe and New Zealand was mainly at species-level, considerable portions of trait data from North America and Australia were at genus or family-level. Assigned traits at family-level may not reflect the real trait diversity within a taxon, for example for the large, ecologically diverse family Chironomidae (\cite{serra_synthesising_2016}). However, species-level identification can be complicated, time-consuming and expensive (\cite{marshall_taxonomic_2006, resh_which_2008}). Consequently, trait information at family-level has been widely used by freshwater ecologists, for example in bioassessment (\cite{beketov_spear_2009}).

@@

\subsection*{Trait aggregation}

When trait values are only available at species-level and observational data are at less precise taxonomic levels, various trait aggregation methods have been used by researchers. We aggregated traits from an Australian and a North American trait dataset to family-level using 5 different aggregation methods and compared the results to traits originally assigned at family-level for these regions. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait for a particular family is. Aggregation of trait information at species or genus-level to a point estimate at family-level suggests a precision that is not necessarily present, especially not for traits with a high variability or if trait information at species-level is missing. Some traits can vary strongly at more precise taxonomic levels than the family-level. For example, \citet{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Further, studies focusing on the lability of traits, i.e. how much traits are constrained by phylogeny, found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and to a lesser extent feeding mode are labile, and thus possibly highly variable (\cite{poff_functional_2006, wilkes_traitbased_2020}). If trait aggregation is necessary, however, our study can give guidance, under which circumstances the choice of aggregation method may be important. The 5 aggregation methods we tested use either the mean or the median and different weightings. Our results indicated that 1) the median aggregation methods compared to mean aggregation methods were often closer to the assigned traits at family-level and 2) the different weightings tested exerted a minor impact on the aggregated trait affinities. We obtained for both datasets more strongly differing cases for mean aggregation methods, irrespective of the weighting. In the Australian dataset, the amount of differing cases was slightly higher for the mean aggregation methods (23 \%) than for median aggregation methods (16 to 18 \% differing cases). In the North American dataset, mean aggregation methods yielded 47 \% cases being different to the assigned traits, a value much higher in comparison to the median aggregation methods (between 15 and 18 \%). When differences occurred between median aggregation methods and assigned traits, these differences were greater compared to the mean aggregation methods, particularly for the North American dataset. There, the mean absolute differences for the median aggregations were twice as high (0.63 - 0.7) compared to the mean aggregations (0.3 - 0.31). These differences can be explained by the binary coding used in the North American trait dataset and by the assigned traits. As a result, traits have affinities of either 1 or 0. Therefore, traits of a particular grouping feature showed in most cases a higher difference in trait affinities to each other than if they were fuzzy coded. Using the median to aggregate binary coded taxa often resulted in a value of 0 or 1, while using the mean yielded values between 0 and 1. Thus, using the median led for the North American dataset either to agreement with the assigned traits or to higher differences compared to the mean aggregation methods.

The weightings explored in this analysis should lead to different aggregated trait affinities in cases where genera within a family have different numbers of species. In fact, the simulation results indicated that for a particular family where one genus has a much larger number of species compared to the other genera, weighting the number of taxa can lead to different aggregated trait affinities, especially for traits with higher levels of variance. However, in the comparison of the trait aggregation methods to assigned traits the number of differing cases and the mean absolute differences in trait affinities were similar across the mean aggregation and across the median aggregation methods, which suggests a small influence of the weightings on the aggregation methods. Also, the distributions of absolute trait affinity differences to assigned traits were relatively similar for the mean aggregation methods and the median aggregation methods, respectively. The minor impact on trait aggregation of the weightings may be explained by the fact that a considerable portion of taxa had low numbers of genera or species. Of the taxa that were compared from the North American trait dataset, 14 \% were identified at family and 62 \% at genus-level, 52 \% comprised 5 or fewer genera, and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset, 21 \% of the compared taxa were identified at family, 40 \% at genus, and 39 \% at species-level, 68 \% of the taxa contained 5 or fewer genera, and 40 \% just one genus (Figure \ref{fig:tax_hierarchy_AUS}). Hence, these results could change when more trait information at species-level becomes available.

>>>>

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Overview_sim_results.png}
  \caption{Ranges of aggregated trait affinities for the three examples of taxonomic hierarchies and simulated levels of trait variability. Shown are the results only for one simulated trait (T1). Similar results were obtained for the other simulated traits (see appendix).  Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $*$ the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots.}
  \label{fig:overview_sim_results}
\end{figure}

@@

Freshwater invertebrate trait databases are a valuable source of data that are freely available to researchers and that can offer opportunities for large-scale, cross-regional ecological study without the need for time-consuming and expensive field work. However, a lack of consistency among regional and continent-wide databases in trait definitions, trait affinity coding, and taxonomic resolution presents challenges to the use of multiple databases for a single study. We explored the effects of harmonising grouping features and aggregating trait affinities at different taxonomic levels from databases from four different continents on the outcomes of trait affinities and on analysis of trait–environment relationships. We found that despite discrepancies in trait definitions and taxonomic resolution across databases, harmonised and aggregated data were useful in identifying trait–environmental relationships, as evidenced by similar results from RDA as were found using observational data. Our results have implications for future multi-regional ecological studies that may benefit from the use of existing databases and provide insight into methods for harmonising traits with differing trait definitions and selecting aggregation methods that minimize either cases or ranges of differences in trait affinity scores when aggregating from higher to lower taxonomic precision.

\subsection{Harmonising trait databases}

Synthesizing trait information from multiple trait databases is crucial for developing the full potential of trait-based approaches, such as for studying community trait responses to environmental gradients across geographic regions. Others have noted the lack of standardised trait terminology in freshwater ecology (\cite{baird_toward_2011, brink_traits-based_2011}), but we are the first to describe invertebrate trait definition discrepancies for some commonly used grouping features. Our experience can offer some insight into the need for, and process of, harmonising invertebrate trait information from trait databases of different regions. Some of the challenges we found were that grouping features were resolved differently into traits in different databases and that various codings (fuzzy, categorical, binary) were used to describe traits. In addition, the same traits were sometimes defined differently, requiring expert knowledge to reclassify these traits (e.g. the trait piercer). Missing information was another challenge. We aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets, but the availability of trait information varied strongly across grouping features and databases. For instance, there was extensive information available for grouping features that are often used in trait–environment studies, such as feeding mode and respiration, but there was surprisingly little information on body form, a trait that is relatively easy to determine.

To resolve definition discrepancies and facilitate data synthesis in the future, terminological standards are needed. Harmonised definitions and concepts of traits have been developed in the past for other organism groups, such as for plants with the \textit{Thesaurus of Plant Characteristics} (TOP) initiative (\cite{garnier_towards_2017}). The core of this harmonisation initiative is to provide standardised trait definitions and to draw connections to synonyms, related terms, and surrounding concepts by linking to other controlled vocabularies or ontologies in the field. Such initiatives could be a role model for freshwater ecologists to establish unambiguous terminologies for invertebrate traits. The existing freshwater invertebrate trait databases could be linked through standardised terminologies or ontologies, as suggested by \citet{baird_toward_2011}. By following the recently proposed \textit{Ecological Trait-data Standard Vocabulary} (ETS), providers of invertebrate trait data could connect their traits to such ontologies (\cite{schneider_towards_2019}). Once standardised trait definitions are established, these will improve trait data sharing, trait data processing when working with multiple trait databases, and, ultimately, interpretation of the derived results.

@@

\subsection*{Trait aggregation method selection}

Trait data for freshwater invertebrates is often resolved to a variety of taxonomic levels because species-level identification can be complicated, time-consuming, and expensive (\cite{marshall_taxonomic_2006, resh_which_2008}). Consequently, trait information at family level has been widely used by freshwater ecologists, for example in bioassessment (\cite{beketov_spear_2009}).
However, assigned traits at family level may not reflect the real trait diversity within, for example, a taxon in the large, ecologically diverse family Chironomidae (\cite{serra_synthesising_2016}). Additionally, various trait aggregation methods have been used by researchers to allow comparison among traits at different taxonomic levels. Evaluation of the differences between aggregated and assigned traits is difficult because it remains unclear what the true value of a particular trait is for a particular family. Aggregation of trait information at species or genus level to a point estimate at family level suggests a precision that is not necessarily present, especially not for traits with high variability or if trait information at species level is missing. Indeed, some traits can vary strongly at more precise taxonomic levels than the family level. For example, \citet{monaghan_improving_2013} found a high intra-family trait dispersion in the Tachet database for traits of the grouping features body size, flow preference, and reproduction. Further, studies focusing on the lability of traits, (i.e. how much traits are constrained by phylogeny), found that traits of ecological preferences (e.g. thermal preference), body size, resistance forms, and, to a lesser extent, feeding mode are labile, and thus possibly highly variable (\cite{poff_functional_2006, wilkes_traitbased_2020}).

If trait aggregation is necessary, however, our study can offer insight about the circumstances under which the choice of aggregation method may be important. The 5 aggregation methods we tested used either the mean or the median and different weightings to assign family-level affinities. Our results indicated that for both the Australian and North American datasets (1) the median aggregation methods provided affinities that were often closer to the assigned traits than were those provided by the mean aggregation methods and (2) the different weighting approaches exerted only a minor influence on the aggregated trait affinities. However, when differences occurred between median aggregation methods and assigned traits, these differences were greater in absolute value compared with the mean aggregation methods, particularly for the North American dataset where mean absolute differences for median aggregations were twice as high as for mean aggregations. This pattern of greater differences for median than mean aggregations can, in large part, be explained by the binary coding used in the North American trait dataset and by the assigned traits. Binary coded traits have affinities of either 1 or 0; therefore, traits of a particular grouping feature showed, in most cases, a higher difference in trait affinities to each other than if they were fuzzy coded. Median-aggregated binary-coded taxa often resulted in a value of 0 or 1, whereas mean aggregation often yielded values between 0 and 1.
Thus, using the median led for the North American dataset either to agreement with the assigned traits or to higher differences compared to the mean aggregation methods. This pattern inherent in using median-aggregation methods to aggregate binary-coded data should be considered when researchers choose methods to resolve affinities at different taxonomic levels.

We expected that, in addition to averaging measure, different weighting approaches to aggregation (i.e., equal weight to each taxon, equal weight to each genus, or weighted by number of species) would affect affinities for families with varying taxonomic hierarchical structure. 
However, we found that although taxonomic structural unevenness appeared to produce variation in affinities among weighting approaches, especially for a family where one genus had a much larger number of species than its other genera, trait variability had a greater effect on the range of affinities (i.e., greater ranges of affinities at higher levels of variance) and on the differences in affinities between weighting methods. Also, the number of differing cases and the mean absolute differences in trait affinities, as well as the distributions of absolute trait affinity differences to assigned traits, were similar across the mean aggregation and across the median aggregation methods, which suggests a small influence of the weighting approach on the aggregation outcomes. The minor impact of the weighting approaches on trait aggregation may be explained by the fact that a considerable portion of taxa had low numbers of genera or species. Of the taxa that were compared from the North American trait dataset, 14 \% were identified at family and 62 \% at genus level, 52 \% comprised five or fewer genera, and 13 \% contained just one genus (Figure \ref{fig:tax_hierarchy_NOA}). In the Australian dataset, 21 \% of the compared taxa were identified at family, 40 \% at genus, and 39 \% at species level, 68 \% of the taxa contained five or fewer genera, and 40 \% just one genus (Figure \ref{fig:tax_hierarchy_AUS}). Hence, these results could change when more species-level trait information becomes available. Our findings show that the choice of aggregation method may matter less when taxonomic hierarchical structure is fairly even and traits are less variable, but the \textit{stepwise\_agg\textsubscript{median}} method tended to produce the widest range in affinities, especially with high taxonomic unevenness and high trait variability.
